On Optimality of Finite Element Integration

### 2016-02-17 TOMS_Paper/assembly.tex
Small fixes here and there...

\begin{abstract}
We tackle the problem of automatically generating optimal finite element integration routines given a high level specification of arbitrary multilinear forms. Global and local optimality are defined in terms of floating point operations given a memory bound. We provide an approach for exploring the space of legal transformations, which systematically reaches a local optimum. The conditions under which optimality is achieved are also discussed. A theoretical analysis and extensive experimentation, which shows significant performance improvements over several state-of-the-art code generation systems, validate the approach.
\end{abstract}

@@

\begin{itemize}
\item We formalize finite element integration loop nests and we build the space of legal transformations impacting their operation count.
\item We provide an algorithm to select points in the transformation space. The algorithm uses a cost model to: (i) understand whether a transformation reduces or increases the operation count; (ii) choose between different (non-composable) transformations.
\item We demonstrate that our algorithm systematically leads to a local optimum. We also explain under what conditions of the input problem global optimality is achieved.
\item We integrate our approach with a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation. The compiler is open-source and available at \url{https://github.com/coneoproject/COFFEE}}, which is in use in the Firedrake framework.
\item We experimentally evaluate using a broader suite of forms, discretizations, and code generation systems than has been used in prior research. This is essential to demonstrate that our optimality model holds in practice.
\end{itemize}

@@

Achieving optimality in polynomial time is not generally feasible, since the $\sigma_{ei}$ sub-expressions can be arbitrarily unstructured. On the other hand, multilinearity suggests a certain degree of regularity for $\alpha_{eij}$ and $\beta_{eik}$. In the following sections, we will elaborate on these observations and formulate an approach that achieves: (i) at least a local optimum in all cases; (ii) optimality whenever the monomials are ``sufficiently structured''. To this purpose, we will construct:
\begin{itemize}
\item the space of legal transformations impacting the operation count (Sections~\ref{sec:sharing-elimination} -- \ref{sec:completeness})
\item an algorithm to select points in the transformation space (Section~\ref{sec:optimal-synthesis})
\end{itemize}

@@

\item Build the \textit{sharing graph} $G= (S, E)$. Each $s \in S$ represents a multilinear symbol or a temporary produced at step (2). An edge $(s^i$, $s^j)$ indicates that a product $s^i s^j$ would appear if the sub-expressions including $s^i$ and $s^j$ were expanded.

@@

\item Partition $S$ into disjoint sets, $\mathbb{S} = \lbrace S^1, ..., S^n\rbrace$, such that $S^i$ includes all instances of a given symbol $s$ in the expression. Transform $G$ by merging $\lbrace s^1, ..., s^m \rbrace \subset S^i$ into a unique vertex $s$ (taking the union of the edges), provided that factorizing $[s^1, ..., s^m]$ would not cause an increase in operation count.\\
\textit{Note: the variation in operation count is a function of (number and arity) of the products wrapping the elements in $S^i$, which impact the expansion cost.}

\item Map $G$ to an Integer Linear Programming (ILP) problem for determining how to optimally apply Strategy~\ref{strategy:ii}. The solution is the set of symbols that will be factorized by Strategy~\ref{strategy:ii}. Let $|S| = n$; the ILP model then is as follows:
\begin{gather*}
x_i \text{: a vertex in } S \text{ (1 if a symbol should be factorized, 0 otherwise)}\\
y_{ij} \text{: an edge in } E \text{ (1 if } s^i \text{ is factorized in the product } s^i s^j \text{ , 0 otherwise)}\\
n_i \text{: the number of edges incident to } x_i \\
\begin{align*}
\min \sum_{i=1}^{n} x_i,\ s.t. ~~~~~~~~~&\sum_{j|(i,j) \in E} y_{ij} \leq n_i x_i,\ i = 1, ..., n \\
& y_{ij} + y_{ji} = 1,\ (i, j) \in E
\end{align*}
\phantom{\hspace{6cm}}
\end{gather*}

@@

After structuring the sharing graph $G$ in such a way that only flop-decreasing transformations are possible, the ILP model is instantiated. At this point, the optimality of the algorithm reduces to prove the correctness of the model, which is relatively straightforward because of its simplicity. The model aims to minimize the operation count by individuating the most promising factorizations. The second set of constraints is to select all edges (i.e., all multiplications), exactly once. The first set of inequalities allows scheduling multiplications: once a vertex $s$ is selected (i.e., once a symbol is decided to be factorized), all multiplications involving $s$ can be grouped. 
\end{enumerate}
\end{proof}

@@

We report the results of our experiments in Figures~\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity} as three-dimensional plots. The axes represent $q$, $nf$, and code transformation system. We show one subplot for each problem instance $\langle form, nf, q\rangle$, with the code transformation system varying within each subplot. The best variant for each problem instance is given by the tallest bar, which indicates the maximum speed-up over non-transformed code. We note that if a bar or a subplot are missing, then the form compiler failed at generating code because of either exceeding the system memory limit or unable to handle the form.

@@

\paragraph{High level view}
Our transformation strategy does not always guarantee minimum execution time. In particular, 5$\%$ of the test cases (3 out of 56, without counting marginal differences) show that \texttt{cfO2} was not optimal in terms of runtime. The most significant of such test cases is the elastic model with $[q=4, nf=0]$. There are two reasons for this. Firstly, low level optimization can have a significant impact on the actual performance. For example, the aggressive loop unrolling in \texttt{tens} eliminates operations on zeros and reduces the working set size by not storing entire temporaries; on the other hand, preserving the loop structure can maximize the chances of autovectorization. Secondly, the transformation strategy adopted when $T_H$ is exceeded plays a key role, as we will later elaborate.

@@

\paragraph{Mass}
We start with the simplest of the bilinear forms investigated, the mass matrix. Results are in Figure~\ref{fig:mass}. We first notice that the lack of improvements when $q=1$ is due to the fact that matrix insertion outweighs local assembly. As $q \geq 2$, \texttt{cfO2} generally shows the highest speed-ups. It is worth noting how \texttt{auto} does not always select the fastest implementation: \texttt{auto} always opts for \texttt{tens}, while as $nf \geq 2$ \texttt{quad} would tend to be preferable. On the other hand, \texttt{cfO2} always makes the optimal decision about whether applying pre-evaluation or not.

\paragraph{Helmholtz}
As happened with the mass matrix problem, when $q=1$ matrix insertion still hides the cost of local assembly. For $q \geq 2$, the general trend is that \texttt{cfO2} outperforms the competitors. In particular, with
\begin{itemize}
\item $nf=0$, the adoption of pre-evaluation by \texttt{cfO2} results in increasingly notable speed-ups over \texttt{cfO1}, as $q$ increases; \texttt{tens} is comparable to \texttt{cfO2}, with \texttt{auto} making the right choice. 
\item $nf=1$, \texttt{auto} picks \texttt{tens}; the choice is however sub-optimal when $q=3$ and $q=4$. This can indirectly be inferred from the large gap between \texttt{cfO1/cfO2} and \texttt{tens/auto}: \texttt{cfO2} applies sharing elimination, but it avoids pre-evaluation. 
\item $nf=2$ and $nf=3$, \texttt{auto} reverts to \texttt{quad}, which would theoretically be the right choice (the flop count is much lower than in \texttt{tens} or what would be produced by pre-evaluation); however, the generated code suffers from the presence of indirection arrays, which break autovectorization and ``traditional'' code motion.
\end{itemize}

The sporadic slow-downs or only marginal improvements exhibited by \texttt{ufls} are imputable to the presence of sharing.

An interesting experiment we performed was relaxing the memory threshold by setting it to $T_H = L3_{size}$. We found that this makes \texttt{cfO2} generally slower as $nf \geq 2$, with a maximum slow-down of 2.16$\times$ with $\langle nf=2, q=2\rangle$. The effects of not having a sensible threshold could even be worse in parallel runs, since the L3 cache is shared by the cores.

\paragraph{Elasticity}
The results for the elastic model are displayed in Figure~\ref{fig:elasticity}. The main observation is that \texttt{cfO2} never triggers pre-evaluation, although in some occasions it should. To clarify this, consider the test case $\langle nf=0, q=2 \rangle$, in which \texttt{tens/auto} show a considerable speed-up over \texttt{cfO2}. \texttt{cfO2} finds pre-evaluation profitable -- that is, actually capable of reducing the operation count -- although it does not apply it because otherwise $T_H$ would be exceeded. However, running the same experiments with $T_H = L3_{size}$ resulted in a dramatic improvement, even higher than that of \texttt{tens}. Our explanation is that despite exceeding $T_H$ by roughly 40$\%$, the save in operation count is so large (5$\times$ in this problem) that pre-evaluation would anyway be the optimal choice. This suggests that our model could be refined to handle the cases in which there is a significant gap between potential cache misses and save in flops.

We also note that:
\begin{itemize}
\item the differences between \texttt{cfO2} and \texttt{cfO1} are due to systematic sharing elimination and the use of symbolic execution to avoid iteration over the zero-valued regions in the basis function tables
\item when $nf=1$, \texttt{auto} prefers \texttt{tens} to \texttt{quad}, which leads to sub-optimal operation counts and execution times
\item \texttt{ufls} generally shows better runtime behaviour than \texttt{quad} and \texttt{tens}. This is due to multiple facts, including avoidance of indirection arrays, preservation of loop structure, a more effective code motion.
\end{itemize}

\paragraph{Hyperelasticity}
In the experiments on the hyperelastic model, shown in Figure~\ref{fig:hyperelasticity}, \texttt{cfO2} exhibits the largest gains out of all problem instances considered in this paper. This is a positive aspect, since it indicates that our transformation algorithm scales well with form complexity. The fact that all code transformation systems (apart from \texttt{tens}) show quite significant speed-ups suggests two points. Firstly, the baseline is highly inefficient. With forms as complex as in the hyperelastic model, a trivial translation of integration routines into code should always be avoided as even the best general-purpose compiler available (Intel's on an Intel platform at maximum optimization level) cannot exploit the structure inherent in the expressions. Secondly, the strategy for removing spatial and temporal sharing has a tremendous impact. Sharing elimination as performed by \texttt{cfO2} ensures a critical reduction in operation count, which becomes particularly pronounced for higher values of $q$.

>>>>

\begin{abstract}
We tackle the problem of automatically generating optimal finite element integration routines given a high level specification of arbitrary multilinear forms. Global and local optimality are defined in terms of floating point operations given a memory bound. We provide an approach for exploring the space of legal transformations, which systematically reaches a local optimum. The conditions under which a global optimum is achieved are also discussed. A theoretical analysis and extensive experimentation, which shows significant performance improvements over several state-of-the-art code generation systems, validate the approach.
\end{abstract}

@@

\begin{itemize}
\item We formalize the class of finite element integration loop nests and we build the space of legal transformations impacting their operation count.
\item We provide an algorithm to select points in the transformation space. The algorithm uses a cost model to: (i) understand whether a transformation reduces or increases the operation count; (ii) choose between different (non-composable) transformations.
\item We demonstrate that our approach systematically leads to a local optimum. We also explain under what conditions of the input problem global optimality is achieved.
\item We integrate our approach with a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation. The compiler is open-source and available at \url{https://github.com/coneoproject/COFFEE}}, which is in use in the Firedrake framework.
\item We experimentally evaluate using a broader suite of forms, discretizations, and code generation systems than has been used in prior research. This is essential to demonstrate that our optimality model holds in practice.
\end{itemize}

@@

Achieving optimality in polynomial time is not generally feasible, since the $\sigma_{ei}$ sub-expressions can be arbitrarily unstructured. On the other hand, multilinearity suggests a certain degree of regularity for $\alpha_{eij}$ and $\beta_{eik}$. In the following sections, we will elaborate on these observations and formulate an approach that achieves: (i) at least a local optimum in all cases; (ii) global optimality whenever the monomials are ``sufficiently structured''. To this purpose, we will construct:
\begin{itemize}
\item the space of legal transformations impacting the operation count (Sections~\ref{sec:sharing-elimination} -- \ref{sec:completeness})
\item an algorithm to select points in the transformation space (Section~\ref{sec:optimal-synthesis})
\end{itemize}

@@

\item Build the \textit{sharing graph} $G= (S, E)$. Each $s \in S$ represents a multilinear symbol or a temporary produced at step (1). An edge $(s^i$, $s^j)$ indicates that a product $s^i s^j$ would appear if the sub-expressions including $s^i$ and $s^j$ were expanded.

@@

\item Partition $S$ into disjoint sets, $\mathbb{S} = \lbrace S^1, ..., S^n\rbrace$, such that $S^i$ includes all instances of a given symbol $s$ in the expression. Transform $G$ by merging $\lbrace s^1, ..., s^m \rbrace \subset S^i$ into a unique vertex $s$ (taking the union of the edges), provided that factorizing $[s^1, ..., s^m]$ will not cause an increase in operation count.\\
\textit{Note: the variation in operation count is a function of (number and arity) of the products wrapping $[s^1, ..., s^m]$, which impact the expansion cost.}

\item Map $G$ to an Integer Linear Programming (ILP) model for determining how to optimally apply Strategy~\ref{strategy:ii}. The solution is the set of symbols that will be factorized by Strategy~\ref{strategy:ii}. Let $|S| = n$; the ILP model then is as follows:
\begin{gather*}
x_i \text{: a vertex in } S \text{ (1 if a symbol should be factorized, 0 otherwise)}\\
y_{ij} \text{: an edge in } E \text{ (1 if } s^i \text{ is factorized in the product } s^i s^j \text{ , 0 otherwise)}\\
n_i \text{: the number of edges incident to } x_i \\
\begin{align*}
\min \sum_{i=1}^{n} x_i,\ s.t. ~~~~~~~~~&\sum_{j|(i,j) \in E} y_{ij} \leq n_i x_i,\ i = 1, ..., n \\
& y_{ij} + y_{ji} = 1,\ (i, j) \in E
\end{align*}
\phantom{\hspace{6cm}}
\end{gather*}
\textit{Note: steps (4) and (5) only impact bilinear forms, since otherwise $E = \emptyset$.}

@@

After structuring the sharing graph $G$ in such a way that only flop-decreasing transformations are possible, the ILP model is instantiated. At this point, proving optimality reduces to convince ourselves about the correctness of the model, which is relatively straightforward because of its simplicity. The model aims to minimize the operation count by individuating the most promising factorizations. The second set of constraints is to select all edges (i.e., all multiplications), exactly once. The first set of inequalities allows scheduling multiplications: once a vertex $s$ is selected (i.e., once a symbol is decided to be factorized), all multiplications involving $s$ can be grouped. 
\end{enumerate}
\end{proof}

@@

We report the results of our experiments in Figures~\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity} as three-dimensional plots. The axes represent $q$, $nf$, and code transformation system. We show one subplot for each problem instance ${\langle} form, nf, q {\rangle}$, with the code transformation system varying within each subplot. The best variant for each problem instance is given by the tallest bar, which indicates the maximum speed-up over non-transformed code. We note that if a bar or a subplot are missing, then the form compiler failed at generating code because of either exceeding the system memory limit or unable to handle the form.

@@

\paragraph{High level view}
Our transformation strategy does not always guarantee minimum execution time. In particular, about 5$\%$ of the test cases (3 out of 56, without counting marginal differences) show that \texttt{cfO2} was not optimal in terms of runtime. The most significant of such test cases is the elastic model with $[q=4,\ nf=0]$. There are two reasons for this. Firstly, low level optimization can have a significant impact on the actual performance. For example, the aggressive loop unrolling in \texttt{tens} eliminates operations on zeros and reduces the working set size by not storing entire temporaries; on the other hand, preserving the loop structure can maximize the chances of autovectorization. Secondly, the transformation strategy adopted when $T_H$ is exceeded plays a key role, as we will later elaborate.

@@

\paragraph{Mass matrix}
We start with the simplest of the bilinear forms investigated, the mass matrix. Results are in Figure~\ref{fig:mass}. We first notice that the lack of improvements when $q=1$ is due to the fact that matrix insertion outweighs local assembly. As $q \geq 2$, \texttt{cfO2} generally shows the highest speed-ups. It is worth noting how \texttt{auto} does not always select the fastest implementation: \texttt{auto} always opts for \texttt{tens}, while as $nf \geq 2$ \texttt{quad} would tend to be preferable. On the other hand, \texttt{cfO2} always makes the optimal decision about whether applying pre-evaluation or not. Surprisingly, despite the simplicity of the form, the performance of the various code generation systems can differ significantly.

\paragraph{Helmholtz}
As in the case of Mass matrix, when $q=1$ the matrix insertion phase is dominant. For $q \geq 2$, the general trend is that \texttt{cfO2} outperforms the competitors. In particular, with
\begin{description}
\item[$nf=0$] pre-evaluation makes \texttt{cfO2} notably faster than \texttt{cfO1}, especially for high values of $q$; \texttt{auto} correctly selects \texttt{tens}, which is comparable to \texttt{cfO2}. 
\item[$nf=1$] \texttt{auto} picks \texttt{tens}; the choice is however sub-optimal when $q=3$ and $q=4$. This can indirectly be inferred from the large gap between \texttt{cfO2} and \texttt{tens/auto}: \texttt{cfO2} applies sharing elimination, but it rightfully avoids pre-evaluation because of the excessive expansion cost. 
\item[$nf=2$ and $nf=3$] \texttt{auto} reverts to \texttt{quad}, which would theoretically be the right choice (the flop count is much lower than in \texttt{tens}); however, the generated code suffers from the presence of indirection arrays, which break autovectorization and ``traditional'' code motion.
\end{description}

The sporadic slow-downs (or marginal improvements) exhibited by \texttt{ufls} are imputable to the presence of sharing in the generated code.

An interesting experiment we additionally performed was relaxing the memory threshold by setting $T_H = L3_{size}$. We found that this makes \texttt{cfO2} generally slower as $nf \geq 2$, with a maximum slow-down of 2.16$\times$ with $\langle nf=2, q=2\rangle$. This effect could be worse when running in parallel, since the L3 cache is shared and different threads would end up competing for the same resource.

\paragraph{Elasticity}
The results for the elastic model are displayed in Figure~\ref{fig:elasticity}. The main observation is that \texttt{cfO2} never triggers pre-evaluation, although in some occasions it should. To clarify this, consider the test case $\langle nf=0,\ q=2 \rangle$, in which \texttt{tens/auto} show a considerable speed-up over \texttt{cfO2}. \texttt{cfO2} finds pre-evaluation profitable in terms of operation count, although it is eventually not applied to avoid exceeding $T_H$. However, running the same experiments with $T_H = L3_{size}$ resulted in a dramatic improvement, even higher than that obtained by \texttt{tens}. The reason is that, despite exceeding $T_H$ by roughly 40$\%$, the save in operation count is so large (5$\times$ in this specific problem) that pre-evaluation would in practice be the winning choice. This suggests that our objective function should be improved to handle the cases in which there is a significant gap between potential cache misses and reduction in operation count.

We also note that:
\begin{itemize}
\item the differences between \texttt{cfO2} and \texttt{cfO1} are due to the perfect sharing elimination and the zero-valued blocks avoidance technique presented in Section~\ref{sec:zeros}.
\item when $nf=1$, \texttt{auto} prefers \texttt{tens} over \texttt{quad}, which leads to sub-optimal operation counts and execution times.
\item \texttt{ufls} often results in better execution times than \texttt{quad} and \texttt{tens}. This is due to multiple facts, including avoidance of indirection arrays, preservation of loop structure, and a more effective code motion strategy.
\end{itemize}

\paragraph{Hyperelasticity}
In the experiments on the hyperelastic model, shown in Figure~\ref{fig:hyperelasticity}, \texttt{cfO2} exhibits the largest gains out of all problem instances considered in this paper. This is a positive aspect, since it indicates that our transformation algorithm scales well with form complexity. The fact that all code transformation systems (apart from \texttt{tens}) show quite significant speed-ups suggests two points. Firstly, the baseline is highly inefficient. With forms as complex as in the hyperelastic model, a trivial translation of integration routines into code should always be avoided as even the best general-purpose compiler available (Intel's on an Intel platform at maximum optimization level) fails at exploiting the structure inherent in the expressions. Secondly, the strategy for removing spatial and temporal sharing has a tremendous impact. Sharing elimination as performed by \texttt{cfO2} ensures a critical reduction in operation count, which becomes particularly pronounced for higher values of $q$.
